---
title: "Making attractive maps in R"
output: html_notebook
---

## Making attractive maps with R and ggplot

### Static maps: using ggplot and colorblind-friendly color schemes

ggplot2 can also be used to make maps, without the ggmap package. I like this because it allows you to plot points (and even polygons!) over a raster layer, and customize how your raster looks, in ways that I think are easier to understand than in base R (e.g., plot() ).. However, there are a few additional data processing steps here, so it's less well-suited for on-the-fly visualization. 

We're also using the [viridis](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html) package for this rockin color scheme. Unlike other color schemes, which often rely on distinctions between green and red, viridis supplies a variety of color schemes that have gradations that are perceptible to folks who can't easily perceive the difference between green and red. 

This uses the gplot() (not ggplot()) command from the rasterVispackage, which is a wrapper around ggplot to use with raster data. A more detailed tutorial for plotting raster data in R is located [here](http://zevross.com/blog/2015/03/30/map-and-analyze-raster-data-in-r/).

This first method requires little data processing, but produces maps that sometimes look striated when I render them to export in markdown. (Not sure why! Something weird with the mapping?) 

Importantly, you'll notice that the spatial resolution of the basemap is only as good as the raster data that you're visualizing.

First, we'll load all the libraries we need. 

```{r}
library(here)
library(tidyverse)
library(raster)
library(ggplot2)

```


Next, we'll choose one of our models to visualize. 

```{r}

#read back in a raster stack of interest
present_In <- stack(here("model_output/rasters/biomod_out.grd"))

#select one layer from the raster stack output
RF_model <- myCurrentProj$Populus.tremuloides_AllData_Full_RF

```


```{r}
# load point data
POTR.dat <- read.csv(POTR.dat, here("data/occurrence/full_POTR.csv"))
```


```{r static maps with geom tile, message = FALSE}

#this is the same plot we created in the previous section

library(rasterVis)
library(viridis)

p = gplot(RF_model) +
  geom_tile(aes(fill = value)) + 
  scale_fill_viridis(na.value = "white") + 
  theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + 
  coord_fixed(ratio = 1.3) # sets the xy resolution to a constant value

p

```

We can use a couple different color schemes in this package. 

```{r}

p + 
  scale_fill_viridis(na.value = "white", option = "magma")

p + 
  scale_fill_viridis(na.value = "white", option = "inferno")

p + 
  scale_fill_viridis(na.value = "white", option = "plasma")


```

We can also plot our initial points over this model. 

```{r}

p + geom_point(POTR.dat %>% arrange(presAbs) %>% filter(presAbs == 1), 
             mapping = aes(x = lon, y = lat, col = factor(presAbs)), shape = 21, color = "white", size = 1) + 
  labs(title = paste0("Aspen occurrence data"))  
  

```

A mapmaking method that requires a little more data prep-- but generates smoother-looking maps on the fly-- is to convert your raster data set to a data frame of points, with a value for each lat-long combination in the data set. This is in contrast to a raster, which is essentially a matrix of data. Confusingly, the geom_raster() function cannot plot an actual raster! 

```{r static maps with geom raster}

#write function to convert raster to ggplot-friendly format
raster_to_pts_gg <- function(raster) {
  spdf <- as(raster, "SpatialPixelsDataFrame")
  df <- as.data.frame(spdf)
  colnames(df) <- c("value", "x", "y")
  df
}

#convert selected raster to list of points
RF_model_pts <- raster_to_pts_gg(RF_model)

#plot
p = ggplot() + 
  geom_raster(data = RF_model_pts, aes(x = x, y = y, fill = value)) + 
  scale_fill_viridis() + 
  #remove default plotting background
  theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + 
  coord_fixed(ratio = 1.3)

p

p + 
  geom_point(POTR.dat %>% arrange(presAbs) %>% filter(presAbs == 1),
             mapping = aes(x = lon, y = lat, col = factor(presAbs)), shape = 21, color = "white", size = 1) + 
  labs(title = paste0("Aspen occurrence data"))  


```

## Leaflet

Mapmaking capabilities in R aren't just limited to static maps-- there are a couple different utilities for making interactive maps in R. The widely-used mapping software [Leaflet](https://leafletjs.com/index.html) can be used in R.

I referred to [this useful demo](https://gbif-europe.github.io/nordic_oikos_2018_r/s3_gbif_demo/gbif_demo.html) for downloading gbif data in R, and then visualizing using leaflet.

First, we need to make sure that the appropriate packages are installed. [mapr](https://cran.r-project.org/web/packages/mapr/index.html) has a simple leaflet mapping utility that we'll install. Conveniently, mapr can take data frames as an input. There is also [Leaflet R package](https://rstudio.github.io/leaflet/).

```{r leaflet in R via mapr }
install.packages("mapr")
library(mapr)

#commented out for rendering to pdf
#visible only in .html format

#map only presences

POTR.dat %>%
  #map_leaflet() needs the species name to be "name"
  mutate(name = "Populus tremuloides") %>%
  #filter to just show presences here
  filter(presAbs == 1) %>%

  map_leaflet(., "lon", "lat", size=2)

```

[link to screenshot of map]

In the interactive map, you can zoom in and out, and mouse over / click on your occurrence points to see data about them. 

Here's a [much more complex Leaflet tutorial](http://strimas.com/r/ebird-county/) that uses specifically eBird data, mapped in R using Leaflet. 

The full leaflet package provides a lot of options to customize your map baselayers, markers, and labels, which are detail in the Leaflet R package tutorial. Without getting too complicated, here's a very simple way to visualize your data in R.

```{r leaflet in R via leaflet}
#install.packages("leaflet")
library(leaflet)

#commented out for rendering to pdf
#visible only in .html format

#filter data to only include presences
#similarly, this takes a lot of time to render when we have this much data.
POTR.dat %>%
  filter(presAbs == 1) %>%
  leaflet() %>%
  addTiles() %>%
  #simple, default markers where you set what shows when you mouse over ("label") and click on ("popup") a point.
  addMarkers(~lon, ~lat, popup = ~as.character(presAbs), label = ~paste0("Presence: ", as.character(presAbs)))

```

[include screenshot of map for pdf]

### Plotting points: using a google-derived basemap

ggmap is a package you can use to download satellite imagery. 

You may need to update other packages on this install. 

```{r}
#install.packages("devtools")
library(devtools)
devtools::install_github("dkahle/ggmap")
library(ggmap)

```

Google provides a way to access baselayers for plotting your points on a map. But, it can throw a lot of errors!

```{r}
library(ggmap)

map = get_map(location = "utah",
              zoom = 6,
              source = "google",
              maptype = "satellite")

p = ggmap(map)
p +
  geom_point(POTR.dat %>% arrange(presAbs),
             mapping = aes(x = lon, y = lat, col = factor(presAbs))) +
  labs(title = paste0("POTR presence and absence"))
```

## Using alternate baselayer sources for maps

An easy way to get around the errors from google maps is just to change the source of the get_map() command. Because google is the default, we have to specify a different source.

### Stamen maps

[Stamen](http://maps.stamen.com/#terrain/12/37.7706/-122.3782) is a data visualization utility provides a few different options for map baselayers-- that are easier to query than google. Similar to the google maps, you're also downloading map tiles from an online utility-- but the stamen maps tend to be less finicky and proprietary.

To use them, use the command get_map(), but set source to "stamen".

You can also read more about the different sources and maptypes in the [documentation for the ggmap package.](https://cran.r-project.org/web/packages/ggmap/ggmap.pdf) 

```{r}
library(ggmap)
map = get_map(location = "Utah",
              zoom = 6,
              source = "stamen")

p = ggmap(map)
p +
  geom_point(POTR.dat %>% arrange(presAbs),
             mapping = aes(x = lon, y = lat, col = factor(presAbs))) +
  labs(title = paste0("POTR presence and absence"))
```




